---
title: "Lab 12 - Smoking during pregnancy"
author: "Anaelle Gackiere"
date: "03-27-2026"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
library(skimr)

# i'm setting my seed right before the bootstrap
```

# Baby Weights

### Exercise 1

Which variables in this dataset are numeric, and what do their distributions look like? Are there any outliers or extreme values that might affect your analysis?

The variables that are numeric in ncbirths are female age, male age, weeks, visits, gained, weight. Mature, premie, lowbirthweight, gender, habit, marital, whitemom are categorical. Gained seems to have extreme values (from 0 pounds to 85 pounds) and it has 27 NAs. fage also has many NAs (171). There are some outliers for a lot of the variables, and we can already tell that smoking status changes some of the distributions.

```{r exercise1-code}

data(ncbirths)

summary(ncbirths)

skimr::skim(ncbirths)

ncbirths %>%
  select(fage, mage, weeks, visits, gained, weight, habit) %>%
  filter(!is.na(habit)) %>%
  pivot_longer(cols = c(fage, mage, weeks, visits, gained, weight),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = habit, y = value, fill = habit)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Numeric Variable Distributions by Smoking Habit",
       x = NULL, y = NULL) +
  theme(legend.position = "none")

```

### Exercise 2

The mean of the weights of the white mom's babies in the dataset is 7.25 pounds.

```{r exercise2-code}

ncbirths_white <- ncbirths %>%
  filter(whitemom == "white")

mean(ncbirths_white$weight, na.rm = TRUE)
```

### Exercise 3

The data is a random sample from NC birth records, so observations should be independent. In terms of sample size, N = 714 is large enough for bootstrapping since the threshold is typically n ≥ 30. The shape of the distribution might pose some concerns; there is no extreme skew (although there is a slight negative skew) but there may be some clustering, since the tail of the distribution is essentially made up of premies.

```{r exercise3-code}

# graphical summaries
ggplot(ncbirths_white, aes(x = weight)) +
  geom_histogram(binwidth = 0.5)

ggplot(ncbirths_white, aes(x = weight, fill = premie)) +
  geom_histogram(binwidth = 0.5, alpha = 0.7)

ggplot(ncbirths_white, aes(y = weight)) +
  geom_boxplot()

# samplesize
nrow(ncbirths_white)
```

### Exercise 4

4a. Use bootstrapping to simulate a distribution of sample means from ncbirths_white.

```{r code-4a}

set.seed(51)

obs_mean <- mean(ncbirths_white$weight, na.rm = TRUE)

boot_means <- ncbirths_white %>%
  specify(response = weight) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")

```

4b. Shift the distribution so that its center aligns with the null hypothesis value (7.43 pounds).

```{r code-4b}

null_dist <- boot_means %>%
  mutate(stat = stat - mean(stat) + 7.43)

```

4c. Create a histogram of your shifted null distribution. Overlay a dashed vertical line to show your observed sample mean.

```{r code-4c}

null_dist %>%
  ggplot(aes(x = stat)) +
  geom_histogram(binwidth = 0.03, fill = "plum", color = "white") +
  geom_vline(xintercept = obs_mean, linetype = "dashed", color = "plum4", linewidth = 1) +
  labs(title = "Null Distribution of Sample Means",
       x = "Mean Birth Weight (lbs)", y = "Count") +
  theme_minimal()



```

4d. Calculate the two-tailed p-value: what proportion of simulated means are at least as extreme (in both directions) as your observed value?

```{r code-4d}

null_dist %>%
  filter(stat <= obs_mean) %>%
  summarize(p_value = 2 * n() / nrow(null_dist))


```

4e. Based on the p-value (0.002) and the visualization, I would conclude that birth weight has changed since 1995.

# Baby weight vs. smoking

### Exercise 5

Both groups have some low weight outliers, and the nonsmoker group has some high weight outliers. Both groups have similar spread, with smokers having a marginally wider IQR compared to nonsmokers. Nonsmokers have a higher median birth weight than smokers. Based on the graph, smokers have a lower SD than nonsmokers.

```{r exercise5-code}

ncbirths %>%
  filter(!is.na(habit)) %>%
  ggplot(aes(x = habit, y = weight, fill = habit)) +
  geom_boxplot() +
  labs(title = "Baby Weight by Mother's Smoking Habit",
       x = "Smoking Habit", y = "Birth Weight (lbs)") +
  theme_minimal() +
  theme(legend.position = "none")


```

### Exercise 6

Removing rows with missing values in either variable ensures that group summaries are based only on complete observations. This prevents functions from returning NA and ensures every observation can be correctly assigned to a group.

```{r exercise6-code}

ncbirths_clean <- ncbirths %>%
  filter(!is.na(habit), !is.na(weight))

```

### Exercise 7

The observed difference in mean birth weight between babies born to smoking and non-smoking mothers is around -.32 lbs.

```{r exercise7-code}

ncbirths_clean %>%
  group_by(habit) %>%
  summarize(mean_weight = mean(weight)) %>%
  summarize(diff = diff(mean_weight))

```

# Mother's age vs. baby weight 
